Graded Call-By-Push-Value

### 2020-07-08 popl21/popl21.tex
[ popl21 ] removed negative variables

Coeffect, quantitative, and many modal type systems maintain a typing
context where each variable $x$ is annotated by a resource qualifier
$r$ in addition to its type $\tau$.  Such a typing context
$(r_1x_1{:}\tau_1, \dots, r_nx_n{:}\tau_n) =: \gamma\Gamma$ can be
split into a pure typing context
$\Gamma = x_1{:}\tau_1, \dots, x_n{:}\tau_n$ and a resource context
$\gamma = x_1{:}r_1, \dots, x_n{:}r_n$ such that
$\dom \gamma = \dom \Gamma$.  We shall freely mix the two notations as
it suits our purpose.

In coeffect-graded CBPV, we wish to interpret each type in the context
as a $\D$-coalgebra for a fixed graded comonad $\D$.  The
interpretation $\den \Gamma$ of the context $\Gamma$ should then be a
comonad coalgebra over the grading $\gamma$ such that judgements
$\qCompTy \gamma \Gamma t N$ can be interpreted as morphisms
$\dent t : \Hom {\Den\Gamma \gamma} {\denn N}$.  Since each variable
comes with its own resource qualifier, we cannot simply model the
context as a tensor product $\tupt {z:\dom\Gamma} \den{\Gamma(z)}$ of
$\D$-coalgebras since this product would indexed by a single resource
qualifier $r \in \R$ rather than a resource context
$\gamma : \dom\Gamma \to \R$.  The solution offered by
\citet{orchard:icfp14} are \emph{structured indexed comonads}.  We
shall generalize this to comonad coalgebras to the extend needed for
interpreting contexts.

@@

  & ::= & \cempty \mid \ext \Gamma x P \mid \graybox{\ext \Gamma y N}

@@

  & ::= & \cempty \mid \ext \gamma x r \mid \ext \gamma y r

@@

  & ::= & \graybox{y} \mid \graybox{\qlet r y u t}
     \mid \qlet {\graybox r} x v t

@@

         \graybox{\qforced r v y t}

@@

 \nru{\rvar^-}{}
     {\ValTy {\gqext 0 \Gamma 1 y N} y N}
 \nru{\rlet^-}
     {\qCompTy \delta \Gamma u N \qquad
      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
    }{\gpCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
 \nru{\rlet^+}

@@

     {\gqValTy \delta \Gamma v {\Box N} \qquad
      \oCompTy {\gqext \gamma \Gamma r y N} t {N'}
    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqforced r v y t} {N'}}

@@

The fundamental novelty is that the typing judgements
$\qValTy \gamma \Gamma v P$ and $\qCompTy \gamma \Gamma t N$ are
equipped with resource contexts $\gamma$ matching the typing contexts
$\Gamma$.  A common pattern in the rules is that resource requirements
of the subterms are added when both subterms are or maybe evaluated at
runtime (rules \rintro\otimes, $\rlet^{+/-}$,
\relim{\Box/\GS/\otimes/\diamond/\lolli}).  Note that in \relim\GS,
the branches $t_i$ of the case statement (\relim\GS) share a resource
context $\gamma$, since only one of the branches is executed at
runtime.  Via rule $\rweak$,
the different resource requirements of the branches can be
subsumed under their maximum wrt.\ $\leq_\R$.  Similarly, the
components $t_i$ of a record (\rintro\Pi) share a resource context
since projection only retrieves one of the components (\relim\Pi).  This is
the opposite of eager tuples (\rintro\otimes) where the elimination
makes all components available at the same time (\relim\otimes), thus,
all components have to be evaluated at runtime.

Function types $\qfun r P N$ are now graded by a resource qualifier
$r$ that specifies how the function argument is to be used in the
function body.  (This type is often written $!_{r}P \lolli N$.)  In
quantitative typing with $\R = \Pot \bN$, qualifier $r$ gives the
possible usage quantities of the function argument, e.g. $\{1\}$ for
exactly one use (linear), $\{0,1\}$ for at most one use (affine),
$\{0\}$ for no use (constant), and $\bN$ for arbitrary use
(unrestricted).  A resource qualifier could also be a security level
(public or private), or a sensitivity level (a non-negative real)
\citep{reedPierce:icfp10}.  If a lambda abstraction $\lam x t$ is
typed with $\qfun r P N$, qualifier $r$ is attached to variable $x$ in
the resource context (rule \rintro\lolli).  If a function
$t : \qfun r P N$ is applied (rule \relim\lolli), we need an
$r$-qualified argument $v : rP$.  We could have given a qualified
value typing judgement $\qValTy \gamma \Gamma v {rP}$ in the style of
\citet{mcBride:wadler60} that---in the quantitative
interpretation---provides $r$ copies of $v$ to be consumed by the
function.  Such a judgement could come with a scaling rule
\[
\ru{\qValTy \gamma \Gamma v {sP}
  }{\qValTy{r\gamma}\Gamma v {(rs)P}}
\]
that allows to scale the production of $v$ by $r$ if the resources are
scaled accordingly (from $\gamma$ to $r\gamma$).  We bake scaling instead
into the transition from values to computations.
Thus, $\relim\lolli$ receives an
argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
the function $\qCompTy \gamma \Gamma t {\qfun r P N}$, the resources
$\delta$ for the argument are scaled by $r$, summing the resource
requirements for the application to $\gamma + r\delta$.
%
Analogous scaling of the eliminatee is baked into the other value
eliminators (\relim\Box, \relim\GS, \relim\otimes), into the
\rlet-rules, and into \rintro\diamond.

@@

As a consequence of eliminatee scaling, $\relim\Box$ is no longer
``natural-deduction'' style like projection, but ``sequent-calculus''
style like $\relim\diamond$.  The original construct $\force v$ could
eliminate an unscaled thunk $\qValTy \gamma \Gamma v {\Box N}$,
however, computation typing $\qCompTy \gamma \Gamma {\force v} N$ has
no scaling option, since negative types $N$ will not be interpreted as
$\D$-coalgebras.  Scaling is semantically restricted to positive
types, and in our type system, to context entries.  Thus, we extend
CBPV to negative variables $y : N$ that come with a resource qualifier
$r$ in the context and can there be interpreted as elements of
$\D[r]\,\denn N$.  This is a modification also taken in call-by-need
CBPV \citep{mcDermottMycroft:esop19}; this compromise allows us to
keep CBPV's polarization of types where positive types stand for
comonad coalgebras and negative types for monad algebras.

@@

Positive types $P$ are interpreted as $\D$-algebras
$\denp P : (\R,\leq) \to \CC$ and negative types $N$ as objects
$\denn N : \CC$, except for in contexts, where they are interpreted as
$\D$-algebras $\Den N r = \D[r]\,\denn N$.  With $\den P = \denp P$,
the interpretation of contexts is thus
$\den \Gamma = \boxtimes_{x:\dom\Gamma} \den{\Gamma(x)}$.
\[
\begin{array}{lcl}
  \Denp\_\_ & : & \PTy \to [(\R,\leq) \to \CC] \\
  \Denp{\Box N}    r & = & \D[r]\, \denn N \\
  \Denp{\sumt I P} r & = & \coprod_{i:I} \Denp{P_i} r \\
  \Denp{\tupt I P} r & = & \otimes_{i:I} \Denp{P_i} r \\
\end{array}
\qquad\qquad
\begin{array}{lcl}
  \denn\_\_ & : & \NTy \to \CC \\
  \denn{\compty r P} & = & \T \Denp P r \\
  \denn{\qfun r P N} & = & \Denp P r \lolli \denn N \\
  \denn{\rect I N}   & = & \prod_{i:I} \denn{N_i} \\
\end{array}
\]
The symmetric monoidal category $\CC$ needs to come with cartesian
products $\rect I B$ as well as with distributive coproducts
$\coprod_I A$ with
$\tdist : \Hom {C \otimes \coprod_I A} {\coprod_{i:I} {(C \otimes A_i)}}$.

Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
$\qValTy \gamma \Gamma v P$ as families
$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
Naturality here means that
$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
    \cdot \gamma)}\right)_{r \leq s}$.
Again, because of $\rweak$, we interpret typing derivations rather than terms.
The interpretation is now rather straightforward, but we spell it out for reference.
\[
\begin{array}{llcl}
\rvar^+
  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
\\
\rintro\Box
  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose[\den \Gamma]
\\
\rintro\GS
  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
  & = & \iota_i \comp \denv v r
\\
\rintro\otimes
  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
\\
\end{array}
\]
\[
\begin{array}{llcl}
\rvar^-
  & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
\\
\rlet^{-}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes
           (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
\\ &&& {~}
  \comp \duplicate
\\
\rlet^{+}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\Box
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\GS
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
  & = & [\dent {t_i}]_{i:I} \comp \tdist
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\otimes
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
  & = & \dent t \comp \isot
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\rintro\diamond
  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
  & = & \treturn \comp \denv v r
\\
\relim\diamond
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
\\ &&& {~}
  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
  \comp \duplicate
\\
\rintro\lolli
  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
  & = & \Lambda \dent t
\\
\relim\lolli
  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
\\
\rintro\Pi
  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
  & = & \langle \dent {t_i} \rangle_{i:I}
\\
\relim\Pi
  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
  & = & \pi_i \comp \dent t
\\
\rweak
  & \dent {\qCompTy \gamma \Gamma t N}
  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
\\
\end{array}
\]

@@

The equational theory of pure CBPV comprises $\beta$ and $\eta$ laws,
and \emph{sequencing laws} that syntactically express that negative
types are monad algebras \citep[Fig.~11]{levi:hosc06}.  The sequencing
laws contain a generalization of the associativity law for monads and
allow to permute a bind under a $\lambda$ or record construction.  In
the presence of $\eta$ for functions and records, they are
inter-derivable with the following permutation laws:
\[
\begin{array}{llcl}
\rpi\diamond
  & \bind {x_2} {(\bind {x_1} {t_1} {t_2})} {t_3}
  & = & \bind {x_1} {t_1} {(\bind {x_2} {t_2} {t_3})}
\\
\rpi\lolli
  & \app {(\bind x u t)} v
  & = & \bind x u {\app t v}
\\
\rpi\Pi
  & \proj i {(\bind x  u t)}
  & = & \bind x u {\proj i t}
\end{array}
\]
Note that permutations for let-bindings like
$\app{(\qlet r x v t)}w = \qlet r x v {\app t w}$ are instances of the
$\beta$-law $\qlet r x v t = \qsubst v r x t$.
%
Herein, \fbox{$\qsubst v r x t$} shall the denote the
(capture-avoiding) replacement $\subst v x t$ of a positive variable
$x$ annotated by $r$ by value $v$ in term $t$.
%
Permutations for value-eliminations like
$\app {(\qsplits r v {\bar x} t)} w = \qsplits r v {\bar x} {\app t
  w}$
are derivable (using the $\beta$-laws) from the $\eta$-laws, like
$\qsubst{v}{r}{z}{t'} = \qsplits r {v} {\bar x} {\qsubst {\ptup {\bar
      x}} r z {t'}}$ for $\otimes$.

Our extension of CBPV features negative variables $y$ standing for
computations $u$, thus, we require a notion of computation
substitution $\qsubst u r t y$ that should have the following semantics
(see corollaries \ref{cor:tysinglesub} and \ref{cor:semsinglesub}).
\begin{align*}
&
  \ru{\qCompTy \delta \Gamma u N \qquad
      \oCompTy {\qext \gamma \Gamma r y N} t {N'}
    }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}}
\\ &
  \dent{\qsubst u r y t} = \dent t
    \comp (\tid[\den\Gamma \gamma]
             \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
    \comp \duplicate[\den\Gamma]
\end{align*}

Further, we add $\beta$-laws for the negative
let-binding and $\relim\Box$, since $\tforce$ now has let-style elimination:
\[
\begin{array}{llcl}
\rbeta{\rlet^{-}}
  & \qlet r y u t
  & = & \qsubst u r y t
\\
\rbeta\Box
  & \qforced r {\thunk u} y t
  & = & \qsubst u r y t
\\
\end{array}
\]
These laws are immediately sound, since left and right hand sides of
these equations have the same denotation.
%
Read as reduction rules, these laws are added to the operational
semantics of CBPV, replacing $\beta$-law $\force{(\thunk u)} = u$.

@@

Substitution for coeffect-graded type system has been worked
out in detail by \citet{atkeyWood:types19,woodAtkey:linearity20}.
It straightforwardly extends to coeffect-graded CBPV.
%
A substitution $\sigma$ is a finite map from variable to terms,
mapping positive variables $x$ to values $v$ and negative variables
$y$ to computations $t$.  Substitution typing
\fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
$\Psi : \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
vector $\Psi(z) : \R^{\dom\Gamma}$ for each variable
$z \in \dom\Delta = \dom \sigma$ in the domain of the substitution.
\[
  \ru{
    }{\SubstTy {0} \Gamma \sempty \cempty}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qValTy \gamma \Gamma v P
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qCompTy \gamma \Gamma t N
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
\]
Capture-avoiding parallel substitution into values \fbox{$v \sigma$} and
computations \fbox{$t \sigma$} is defined as usual by recursion on the term.

>>>>

Coeffect, quantitative, and many modal type systems maintain a typing
context where each variable $x$ is annotated by a resource qualifier
$r$ in addition to its type $P$.  Such a typing context
$(r_1x_1{:}P_1, \dots, r_nx_n{:}P_n) =: \gamma\Gamma$ can be
split into a pure typing context
$\Gamma = x_1{:}P_1, \dots, x_n{:}P_n$ and a resource context
$\gamma = x_1{:}r_1, \dots, x_n{:}r_n$ such that
$\dom \gamma = \dom \Gamma$.  We shall freely mix the two notations as
it suits our purpose.

In coeffect-graded CBPV, we wish to interpret each type in the context
as a $\D$-coalgebra for a fixed graded comonad $\D$.  The
interpretation $\den \Gamma$ of the context $\Gamma$ should then be a
comonad coalgebra over the grading $\gamma$ such that judgements
$\qCompTy \gamma \Gamma t N$ can be interpreted as morphisms
$\dent t : \Hom {\Den\Gamma \gamma} {\denn N}$.  Since each variable
comes with its own resource qualifier, we cannot simply model the
context as a tensor product $\tupt {x:\dom\Gamma} \den{\Gamma(x)}$ of
$\D$-coalgebras since this product would indexed by a single resource
qualifier $r \in \R$ rather than a resource context
$\gamma : \dom\Gamma \to \R$.  The solution offered by
\citet{orchard:icfp14} are \emph{structured indexed comonads}.  We
shall generalize this to comonad coalgebras to the extend needed for
interpreting contexts.

@@

  & ::= & \cempty \mid \ext \Gamma x P % \mid \graybox{\ext \Gamma y N}

@@

  & ::= & \cempty \mid \ext \gamma x r % \mid \ext \gamma y r

@@

  & ::= &
     %     \graybox{y} \mid \graybox{\qlet r y u t}
     % \mid
         \qlet {\graybox r} x v t

@@

         \force v
         % \graybox{\qforced r v y t}

@@

% \graybox{
%  \nru{\rvar^-}{}
%      {\ValTy {\gqext 0 \Gamma 1 y N} y N}
% }
% \graybox{
%  \nru{\rlet^-}
%      {\qCompTy \delta \Gamma u N \qquad
%       \oCompTy {\qext \gamma \Gamma r y N} t {N'}
%     }{\gpCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
% }
% \\[1.5ex]
%  \nru{\rlet^+}
 \nru{\rlet}

@@

     {\qValTy \gamma \Gamma v {\Box N}
    }{\qCompTy \gamma \Gamma {\force v} {N}}
 % \nru{\relim\Box}
 %     {\gqValTy \delta \Gamma v {\Box N} \qquad
 %      \oCompTy {\gqext \gamma \Gamma r y N} t {N'}
 %    }{\gpCompTy {\gamma + r\delta} \Gamma {\gqforced r v y t} {N'}}

@@

The fundamental novelty is that the typing judgements
$\qValTy \gamma \Gamma v P$ and $\qCompTy \gamma \Gamma t N$ are
equipped with resource contexts $\gamma$ matching the typing contexts
$\Gamma$.  A common pattern in the rules is that resource requirements
of the subterms are added when both subterms are or maybe evaluated at
runtime (rules \rintro\otimes, \rlet, %$\rlet^{+/-}$,
\relim{%\Box/
\GS/\otimes/\diamond/\lolli}).  Note that in \relim\GS,
the branches $t_i$ of the case statement (\relim\GS) share a resource
context $\gamma$, since only one of the branches is executed at
runtime.  Via rule $\rweak$,
the different resource requirements of the branches can be
subsumed under their maximum wrt.\ $\leq_\R$.  Similarly, the
components $t_i$ of a record (\rintro\Pi) share a resource context
since projection only retrieves one of the components (\relim\Pi).  This is
the opposite of eager tuples (\rintro\otimes) where the elimination
makes all components available at the same time (\relim\otimes), thus,
all components have to be evaluated at runtime.

Function types $\qfun r P N$ are now graded by a resource qualifier
$r$ that specifies how the function argument is to be used in the
function body.  (This type is often written $!_{r}P \lolli N$.)  In
quantitative typing with $\R = \Pot \bN$, qualifier $r$ gives the
possible usage quantities of the function argument, e.g. $\{1\}$ for
exactly one use (linear), $\{0,1\}$ for at most one use (affine),
$\{0\}$ for no use (constant), and $\bN$ for arbitrary use
(unrestricted).  A resource qualifier could also be a security level
(public or private), or a sensitivity level (a non-negative real)
\citep{reedPierce:icfp10}.  If a lambda abstraction $\lam x t$ is
typed with $\qfun r P N$, qualifier $r$ is attached to variable $x$ in
the resource context (rule \rintro\lolli).  If a function
$t : \qfun r P N$ is applied (rule \relim\lolli), we need an
$r$-qualified argument $v : rP$.  We could have given a qualified
value typing judgement $\qValTy \gamma \Gamma v {rP}$ in the style of
\citet{mcBride:wadler60} that---in the quantitative
interpretation---provides $r$ copies of $v$ to be consumed by the
function.  Such a judgement could come with a scaling rule
\[
\ru{\qValTy \gamma \Gamma v {sP}
  }{\qValTy{r\gamma}\Gamma v {(rs)P}}
\]
that allows to scale the production of $v$ by $r$ if the resources are
scaled accordingly (from $\gamma$ to $r\gamma$).
%
However, this would have prevented the typing of $\force v$ (rule
\relim\Box), bcause there is no place for the scaling factor in
computation typing.  Semantically, $v$ lives in $\D[\_]\denn N$, and
we can only $\textract$ the computation in $\denn N$ if we can
instantiate the coeffect qualifier $\_$ to $1$.  Such is not possible
if scaling already happened (and needs to be respected).
%
Instead, we bake scaling
into the transition from values to computations.
Thus, $\relim\lolli$ receives an
argument $\qValTy \delta \Gamma v P$, and to satisfy the demands of
the function $\qCompTy \gamma \Gamma t {\qfun r P N}$, the resources
$\delta$ for the argument are scaled by $r$, summing the resource
requirements for the application to $\gamma + r\delta$.
%
Analogous scaling of the eliminatee is baked into the other value
eliminators (%\relim\Box,
\relim\GS, \relim\otimes), into %the \rlet-rules,
 \rlet,
and into \rintro\diamond.

@@

% As a consequence of eliminatee scaling, $\relim\Box$ is no longer
% ``natural-deduction'' style like projection, but ``sequent-calculus''
% style like $\relim\diamond$.  The original construct $\force v$ could
% eliminate an unscaled thunk $\qValTy \gamma \Gamma v {\Box N}$,
% however, computation typing $\qCompTy \gamma \Gamma {\force v} N$ has
% no scaling option, since negative types $N$ will not be interpreted as
% $\D$-coalgebras.  Scaling is semantically restricted to positive
% types, and in our type system, to context entries.  Thus, we extend
% CBPV to negative variables $y : N$ that come with a resource qualifier
% $r$ in the context and can there be interpreted as elements of
% $\D[r]\,\denn N$.  This is a modification also taken in call-by-need
% CBPV \citep{mcDermottMycroft:esop19}; this compromise allows us to
% keep CBPV's polarization of types where positive types stand for
% comonad coalgebras and negative types for monad algebras.

@@

Positive types $P$ are interpreted as $\D$-algebras
$\denp P : (\R,\leq) \to \CC$ and negative types $N$ as objects
$\denn N : \CC$.
% , except for in contexts, where they are interpreted as
% $\D$-algebras $\Den N r = \D[r]\,\denn N$.  With $\den P = \denp P$,
% the interpretation of contexts is thus
% $\den \Gamma = \boxtimes_{x:\dom\Gamma} \den{\Gamma(x)}$.
The interpretation of contexts is
$\den \Gamma = \boxtimes_{x:\dom\Gamma} \denp{\Gamma(x)}$.
\[
\begin{array}{lcl}
  \Denp\_\_ & : & \PTy \to [(\R,\leq) \to \CC] \\
  \Denp{\Box N}    r & = & \D[r]\, \denn N \\
  \Denp{\sumt I P} r & = & \coprod_{i:I} \Denp{P_i} r \\
  \Denp{\tupt I P} r & = & \otimes_{i:I} \Denp{P_i} r \\
\end{array}
\qquad\qquad
\begin{array}{lcl}
  \denn\_\_ & : & \NTy \to \CC \\
  \denn{\compty r P} & = & \T \Denp P r \\
  \denn{\qfun r P N} & = & \Denp P r \lolli \denn N \\
  \denn{\rect I N}   & = & \prod_{i:I} \denn{N_i} \\
\end{array}
\]
The symmetric monoidal category $\CC$ needs to be equipped with cartesian
products $\rect I B$ as well as with distributive coproducts
$\coprod_I A$ with
$\tdist : \Hom {C \otimes \coprod_I A} {\coprod_{i:I} {(C \otimes A_i)}}$.

Computations $\qCompTy \gamma \Gamma t N$ are interpreted as morphisms
$\dent t : \Hom {\Den\Gamma\gamma} {\denn N}$ and values
$\qValTy \gamma \Gamma v P$ as families
$\denv v r : \Hom {\Den\Gamma{r\gamma}} {\Denp P r}$ natural in $r$.
Naturality here means that
$\Denp P {r \leq s} \comp \denv v r = \denv v s \comp \left(\Den \Gamma {(\_
    \cdot \gamma)}\right)_{r \leq s}$.
Again, because of $\rweak$, we interpret typing derivations rather than terms.
The interpretation is now rather straightforward, but we spell it out for reference.
\[
\begin{array}{llcl}
\rvar^+
  & \denv {\ValTy{\qext 0 \Gamma 1 x P} x P} r
  & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \tid[\Denp P r])
\\
\rintro\Box
  & \denv {\qValTy \gamma \Gamma {\thunk t} {\Box N}} r
  & = & \fmapt[{\D[r]}]{\dent t} \comp \expose[\den \Gamma]
\\
\rintro\GS
  & \denv{\qValTy \gamma \Gamma {\inj i v} {\sumt I P}} r
  & = & \iota_i \comp \denv v r
\\
\rintro\otimes
  & \denv{\pValTy {\sumt I \gamma} \Gamma {\tup v} {\tupt I P}} r
  & = & \tupt {i:I} \denv {v_i}r \comp \duplicate
\\
\end{array}
\]
\[
\begin{array}{llcl}
% \rvar^-
%   & \dent {\oCompTy{\qext 0 \Gamma 1 y N} y N}
%   & = & \Gr \comp (\tdrop[\den\Gamma] \otimes \extract[\denn N])
% \\
% \rlet^{-}
%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r y u t} {N'}}
%   & = & \dent t
%   \comp (\tid[\Den \Gamma \gamma] \otimes
%            (\fmapt[{\D[r]}] \dent u \comp \expose[\den \Gamma]))
% \\ &&& {~}
%   \comp \duplicate
% \\
\rlet %^{+}
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qlet r x v t} {N}}
  & = & \dent t
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\Box
  & \dent {\qCompTy \gamma \Gamma {\force v} {N}}
  & = & \extract[\D] \comp \denv v 1
% \relim\Box
%   & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qforced r v y t} {N'}}
%   & = & \dent t
%   \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
%   \comp \duplicate
\\
\relim\GS
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\caser v \{r x_i.\, t_i\}_{i:I}} {N}}
  & = & [\dent {t_i}]_{i:I} \comp \tdist
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\relim\otimes
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qsplits r v {\bar x} t} {N}}
  & = & \dent t \comp \isot
  \comp (\tid[\Den \Gamma \gamma] \otimes \denv v r)
  \comp \duplicate
\\
\rintro\diamond
  & \dent {\pCompTy {r\gamma} \Gamma {\ret v} {\compty r P}}
  & = & \treturn \comp \denv v r
\\
\relim\diamond
  & \dent {\pCompTy {\gamma + r\delta} \Gamma {\qbind r x u t} {N}}
  & = & \run[\den N] \comp \fmapt[\T]{\dent t} \comp \strengthl
\\ &&& {~}
  \comp (\tid[\Den \Gamma \gamma] \otimes \dent u)
  \comp \duplicate
\\
\rintro\lolli
  & \dent {\qCompTy \gamma \Gamma {\lam x t} {\qfun r P N}}
  & = & \Lambda \dent t
\\
\relim\lolli
  & \dent{\pCompTy {\gamma + r\delta} \Gamma {\app t v} N}
  & = & \teval \comp (\dent t \otimes \denv v r) \comp \duplicate
\\
\rintro\Pi
  & \dent {\qCompTy \gamma \Gamma {\recrd I t} {\rect I N}}
  & = & \langle \dent {t_i} \rangle_{i:I}
\\
\relim\Pi
  & \dent {\qCompTy \gamma \Gamma {\proj i t} {N_i}}
  & = & \pi_i \comp \dent t
\\
\rweak
  & \dent {\qCompTy \gamma \Gamma t N}
  & = & \dent t \comp \Den \Gamma {\gamma \leq \gamma'}
\\
\end{array}
\]

@@

The equational theory of pure CBPV comprises $\beta$ and $\eta$ laws,
and \emph{sequencing laws} that syntactically express that negative
types are monad algebras \citep[Fig.~11]{levy:hosc06}.  The sequencing
laws contain a generalization of the associativity law for monads and
allow to permute a bind under a $\lambda$ or record construction.  In
the presence of $\eta$ for functions and records, they are
inter-derivable with the following permutation laws:
\[
\begin{array}{llcl}
\rpi\diamond
  & \bind {x_2} {(\bind {x_1} {t_1} {t_2})} {t_3}
  & = & \bind {x_1} {t_1} {(\bind {x_2} {t_2} {t_3})}
\\
\rpi\lolli
  & \app {(\bind x u t)} v
  & = & \bind x u {\app t v}
\\
\rpi\Pi
  & \proj i {(\bind x  u t)}
  & = & \bind x u {\proj i t}
\end{array}
\]
Note that permutations for let-bindings like
$\app{(\qlet r x v t)}w = \qlet r x v {\app t w}$ are instances of the
$\beta$-law $\qlet r x v t = \qsubst v r x t$.
%
Herein, \fbox{$\qsubst v r x t$} shall the denote the
(capture-avoiding) replacement $\subst v x t$ of a positive variable
$x$ annotated by $r$ by value $v$ in term $t$.
%
Permutations for value-eliminations like
$\app {(\qsplits r v {\bar x} t)} w = \qsplits r v {\bar x} {\app t
  w}$
are derivable (using the $\beta$-laws) from the $\eta$-laws, like
$\qsubst{v}{r}{z}{t'} = \qsplits r {v} {\bar x} {\qsubst {\ptup {\bar
      x}} r z {t'}}$ for $\otimes$.

% Our extension of CBPV features negative variables $y$ standing for
% computations $u$, thus, we require a notion of computation
% substitution $\qsubst u r t y$ that should have the following semantics
% (see corollaries \ref{cor:tysinglesub} and \ref{cor:semsinglesub}).
% \begin{align*}
% &
%   \ru{\qCompTy \delta \Gamma u N \qquad
%       \oCompTy {\qext \gamma \Gamma r y N} t {N'}
%     }{\pCompTy {\gamma + r\delta} \Gamma {\qsubst u r y t} {N'}}
% \\ &
%   \dent{\qsubst u r y t} = \dent t
%     \comp (\tid[\den\Gamma \gamma]
%              \otimes (\fmapt[{\D[r]}] \dent u \comp \expose[\den\Gamma]))
%     \comp \duplicate[\den\Gamma]
% \end{align*}

% Further, we add $\beta$-laws for the negative
% let-binding and $\relim\Box$, since $\tforce$ now has let-style elimination:
% \[
% \begin{array}{llcl}
% \rbeta{\rlet^{-}}
%   & \qlet r y u t
%   & = & \qsubst u r y t
% \\
% \rbeta\Box
%   & \qforced r {(\thunk u)} y t
%   & = & \qsubst u r y t
% \\
% \end{array}
% \]
% These laws are immediately sound, since left and right hand sides of
% these equations have the same denotation.
% %
% Read as reduction rules, these laws are added to the operational
% semantics of CBPV, replacing $\beta$-law $\force{(\thunk u)} = u$.

@@

Substitution for coeffect-graded type system has been worked
out in detail by \citet{atkeyWood:types19,woodAtkey:linearity20}.
It straightforwardly extends to coeffect-graded CBPV.
%
A substitution $\sigma$ is a finite map from variable to terms,
mapping positive variables $x$ to values $v$ and negative variables
$y$ to computations $t$.  Substitution typing
\fbox{$\SubstTy \Psi \Gamma \sigma \Delta$} is equipped with a matrix
$\Psi : \R^{\dom\Gamma \times \dom\Delta}$ recording the usage
vector $\Psi(z) : \R^{\dom\Gamma}$ for each variable
$z \in \dom\Delta = \dom \sigma$ in the domain of the substitution.
\[
  \ru{
    }{\SubstTy {0} \Gamma \sempty \cempty}
\qquad
  \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
      \qValTy \gamma \Gamma v P
    }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma x v} {\ext \Delta x P}}
% \qquad
%   \ru{\SubstTy \Psi \Gamma \sigma \Delta \qquad
%       \qCompTy \gamma \Gamma t N
%     }{\pSubstTy {\Psi,\gamma} \Gamma {\sext \sigma y t} {\ext \Delta y N}}
\]
Capture-avoiding parallel substitution into values \fbox{$v \sigma$} and
computations \fbox{$t \sigma$} is defined as usual by recursion on the term.
